---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: yes
---
## Reproducible Research Course Project 1
Details can be found [here](https://www.coursera.org/learn/reproducible-research/peer/gYyPt/course-project-1).

## Loading and preprocessing the data
We load the data directly from the zip archive "activity.zip" and transform the "date" column to an appropriate date format.
```{r message = FALSE}
#use message = FALSE to hide messages from loading the library
library(dplyr)
act <- read.table(unz("activity.zip","activity.csv"), header = TRUE, sep = ",")
act <- transform(act, date = as.Date(date))
```

## What is mean total number of steps taken per day?
```{r message = FALSE}
#use message = FALSE to hide messages from loading the library
library(ggplot2)
total_steps <- act %>% group_by(date) %>% summarize(steps = sum(steps, na.rm = TRUE))
qplot(steps, data = total_steps, bins = 50, main = "Daily steps", xlab = "Number of steps per day", ylab = "Count")

mean_steps <- mean(total_steps$steps)
median_steps <- median(total_steps$steps)
```

The mean of the total steps per day is `r mean_steps` and the median is `r median_steps`.

## What is the average daily activity pattern?

```{r}
avg_steps <- act %>%  group_by(interval) %>% summarize(steps = mean(steps, na.rm = TRUE))
qplot(interval, steps, data=avg_steps, geom = "line", main = "Steps per time interval", xlab = "Interval", ylab = "Steps")
```

```{r}
max_index <- which(avg_steps$steps == max(avg_steps$steps))
max_interval <- avg_steps$interval[max_index]
```

The average maximum of steps occur in the interval `r max_interval`.

## Imputing missing values
### Number of missing values
```{r}
#there are no missing dates or intervals, it suffices to calculate the missing steps
missing_values <- sum(is.na(is.na(act$steps)))
```

The number of rows with missing values is `r missing_values`.

### Imputation
We impute the missing steps by taking the mean of the respective time interval.
```{r}
#impute values by taking the mean by interval
imputed_values <- act %>% group_by(interval) %>% summarize(mean = mean(steps, na.rm = TRUE))

#create a copy
act_impute <- data.frame(act)
#fill in the missing steps with the imputed values
for(i in 1:nrow(act_impute)){
    if(is.na(act_impute$steps[i])){
        index <- which(imputed_values$interval == act_impute$interval[i])
        act_impute$steps[i] <- imputed_values$mean[index]
    }
}

```

Next, we take a look at the average daily activity pattern again.

```{r}
total_steps_imputed <- act_impute %>% group_by(date) %>% summarize(steps = sum(steps))
qplot(steps, data = total_steps_imputed, bins = 50, main = "Daily Steps", xlab = "Number of steps per day", ylab = "Count")

imputed_mean_steps <- mean(total_steps_imputed$steps)
imputed_median_steps <- median(total_steps_imputed$steps)
```

When missing steps are imputed as above, the mean of the total steps per day is `r format(imputed_mean_steps, scientific=FALSE)` and the median is `r format(imputed_median_steps, scientific=FALSE)`. While the mean and median do not change too much, values around the mean become much more prominent after imputation.

## Are there differences in activity patterns between weekdays and weekends?
We add a column indicating wether there respective date is a weekday or not. For this step, we use the imputed values from before.
```{r}
#transform weekdays to numbers (1 == Monday)
wd <- as.numeric(strftime(act$date, "%u"))

#helper function to turn number of the day into "weekday" and "weekend"
weekday_or_weekend <- function(row){
    if(row < 6){
        return("weekday")
    } else{
        return("weekend")    
    }
}

#create new weekend/weekday column
act_impute$day <- as.factor(sapply(wd, weekday_or_weekend))
```
### Activity patterns of weekdays and weekends
```{r}
avg_steps_day <- act_impute %>%  group_by(day, interval) %>% summarize(steps = mean(steps, na.rm = TRUE))

g <- qplot(interval, steps, data=avg_steps_day, geom = "line", main = "Steps per time interval per kind of day", xlab = "Interval", ylab = "Number of steps") 
g <- g + facet_grid(rows = day ~ .)
print(g)
```